---
title: "Project Milestone 2"
author: "Peter Ward, Hayes Waddell, Nathan Gumagay"
format: 
  html:
    embed-resources: true
---

# Read in Data and Data Cleaning

```{r}
# Reading in our player name and player data datasets

library(readr)
library(tidyverse)
# Data from first half of 2024 season for every player
baseballdata <- read_csv("../data/swing_data.csv")

# Pitcher name data
playerdata <- read_csv("../data/player_data.csv")

# Data from entire 2024 season, only containing qualified pitchers
pitcherdata <- read_csv("../data/pitcher_data.csv")

# Choosing relevant variables from name dataset
playerdataclean <- playerdata %>%
  select(PLAYERNAME, LASTNAME, MLBID, THROWS)
```

```{r}
# Data cleaning including selecting relevant variables that might be useful in our analysis, joining datasets to include pitcher's names in our main dataset, and filtering out any data points that are NA for bat_speed and swing_length (since we will rely heavily on these variables for our analysis), as well as any data points that don't involve any swing being made(takes, hit by pitches, bunts, etc.) 

baseballdataclean <- baseballdata %>%
  select(game_date, release_speed, player_name, events, zone, stand, bb_type, balls, strikes, pfx_x, pfx_z, outs_when_up, inning, hit_distance_sc, launch_angle, launch_speed, effective_speed, release_spin_rate, launch_speed_angle, estimated_ba_using_speedangle, estimated_woba_using_speedangle, woba_value, bat_score, fld_score, pitch_name, pitcher, bat_speed, swing_length, description)

baseballdataclean2 <- baseballdataclean %>% 
  left_join(playerdataclean, by = c("pitcher" = "MLBID"))

baseballdataclean2 <- baseballdataclean2 %>%
  mutate(pitch_name = recode(pitch_name, "4-Seam Fastball" = "4 Seam")) %>%
  mutate(pitch_name = recode(pitch_name, "Knuckle Curve" = "K Curve")) %>%
  mutate(pitch_name = recode(pitch_name, "Split-Finger" = "Splitter"))

baseballdataclean2 <- baseballdataclean2 %>%
  filter(bat_speed != 'NA') %>%
  filter(swing_length != 'NA') %>%
  filter(!description %in% c("ball", "called_strike", "blocked_ball", "pitchout", "hit_by_pitch", "foul_bunt","missed_bunt", "bunt_foul_tip"))

baseballdataclean2 <- baseballdataclean2 %>%
  filter(launch_speed != 'NA') %>%
  filter(PLAYERNAME != 'NA') %>%
  mutate(teamwinning = ifelse(bat_score > fld_score, TRUE, FALSE)) %>%
  mutate(sameside = ifelse(THROWS == stand, TRUE, FALSE)) %>%
  group_by(PLAYERNAME) %>%
  mutate(tot = n())

baseballdatacleanqual <- baseballdataclean2 %>%
  filter(tot > 50) %>%
  filter(!events %in% c("sac_bunt", "catcher_interf")) %>%
  mutate(speeddiff = effective_speed - release_speed)
  
```

```{r}
baseballdataclean2 %>%
  filter(launch_speed != 'NA') %>%
  ggplot(aes(x =  bat_speed, y = launch_speed)) +
  geom_point() +
  geom_smooth(method = 'lm', se = TRUE) +
  labs(title = "Relationship Between Bat Speed and Exit Velocity", y = "Exit Velocity") +
  theme_minimal()
```

```{r}
baseballdatacleanqual %>%
  filter(pitch_name != 'Slow Curve' & pitch_name != 'Forkball') %>%
  ggplot(aes(x = pitch_name, y = bat_speed, fill = pitch_name)) +
  geom_boxplot() +
  theme_minimal() +
  labs(x = "Pitch Name", 
       y = "Swing Speed", 
       title = "Distribution of Batters' Swing Speeds by Pitch Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```




# THE FOLLOWING ANALYSIS USES THE DATASET FROM THE FIRST HALF OF THE 2024 SEASON

```{r}
meanbatspeeds <- baseballdatacleanqual %>%
  group_by(player_name) %>%
  summarize(meanbatspeed = mean(bat_speed, na.rm = TRUE))

pitchermeanbatspeeds <- baseballdatacleanqual %>%
  group_by(PLAYERNAME) %>%
  summarize(meanbatspeedagainst = mean(bat_speed, na.rm = TRUE))

baseballdatacleanqual <- baseballdatacleanqual %>%
  left_join(meanbatspeeds)

baseballdatacleanqual <- baseballdatacleanqual %>%
  mutate(batspeeddiff = bat_speed - meanbatspeed)

baseballdatacleanqual <- baseballdatacleanqual %>%
  mutate(fasterthanavg = if_else(bat_speed > meanbatspeed, TRUE, FALSE))

fastavgagainst <- baseballdatacleanqual %>%
  group_by(PLAYERNAME) %>%
  summarize(avgbatspeeddiff = mean(batspeeddiff, na.rm = TRUE)) %>%
  arrange(desc(avgbatspeeddiff)) %>%
  head(25)

slowavgagainst <- baseballdatacleanqual %>%
  group_by(PLAYERNAME) %>%
  summarize(avgbatspeeddiff = mean(batspeeddiff, na.rm = TRUE)) %>%
  arrange(avgbatspeeddiff) %>%
  head(25)
```
```{r}
fastavgagainstall <- baseballdatacleanqual %>%
  inner_join(fastavgagainst, by = 'PLAYERNAME')

slowavgagainstall <- baseballdatacleanqual %>%
  inner_join(slowavgagainst, by = 'PLAYERNAME')

fastvelo <- fastavgagainstall %>%
  group_by(pitch_name) %>%
  summarize(meanvelofast = mean(release_speed, na.rm = TRUE))

slowvelo <- slowavgagainstall %>%
  group_by(pitch_name) %>%
  summarize(meanveloslow = mean(release_speed, na.rm = TRUE))

allvelo <- fastvelo %>%
  full_join(slowvelo)

allvelo <- allvelo %>%
  mutate(velodiff = meanveloslow - meanvelofast) %>%
  rename(slow = meanveloslow, fast = meanvelofast)

allvelo2 <- allvelo %>%
  pivot_longer(cols = c(slow, fast), 
               names_to = "BatSpeedAgainst", 
               values_to = "Velocity")

allvelo2 %>%
  ggplot(aes(x = pitch_name, y = Velocity, fill = BatSpeedAgainst)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(x = "Pitch Type", y = "Velocity (mph)", title = "Mean Velocity Comparison by Pitch Type", fill = "Bat Speed Against", caption = "Takeaway: Throwing 4 Seam Fastballs Harder Leads to Slower Bat Speeds Against")
```

```{r}
fastspinrate <- fastavgagainstall %>%
  group_by(pitch_name) %>%
  summarize(meanspinfast = mean(release_spin_rate, na.rm = TRUE))

slowspinrate <- slowavgagainstall %>%
  group_by(pitch_name) %>%
  summarize(meanspinslow = mean(release_spin_rate, na.rm = TRUE))

allspinrate <- fastspinrate %>%
  full_join(slowspinrate)

allspinrate<- allspinrate %>%
  rename(slow = meanspinslow, fast = meanspinfast)

allspinrate2 <- allspinrate %>%
  pivot_longer(cols = c(slow, fast), 
               names_to = "BatSpeedAgainst", 
               values_to = "SpinRate")

allspinrate2 %>%
  ggplot(aes(x = pitch_name, y = SpinRate, fill = BatSpeedAgainst)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(x = "Pitch Type", y = "Spin Rate (rpm)", title = "Mean Spin Rate Comparison by Pitch Type", fill = "Bat Speed Against", caption = "Takeaway: Pitchers With Slow Bat Speed Against: Less spin on curveballs, more spin on horizontally breaking pitches")
```
```{r}
fastspeeddiff <- fastavgagainstall %>%
  group_by(pitch_name) %>%
  summarize(meanspeeddiff_fast = mean(speeddiff, na.rm = TRUE))

slowspeeddiff <- slowavgagainstall %>%
  group_by(pitch_name) %>%
  summarize(meanspeeddiff_slow = mean(speeddiff, na.rm = TRUE))

allspeeddiff <- fastspeeddiff %>%
  full_join(slowspeeddiff)

allspeeddiff <- allspeeddiff %>%
  rename(slow = meanspeeddiff_slow, fast = meanspeeddiff_fast)

allspeeddiff2<- allspeeddiff %>%
  pivot_longer(cols = c(slow, fast), 
               names_to = "BatSpeedAgainst", 
               values_to = "SpeedDifference")

allspeeddiff2 %>%
  ggplot(aes(x = pitch_name, y = SpeedDifference, fill = BatSpeedAgainst)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(x = "Pitch Type", y = "Mean Perceived Velocity - Actual Velocity (mph)", title = "Mean Perceived vs Actual Pitch Speed Comparison by Pitch Type", fill = "Bat Speed Against", caption = "Takeaway: Pitchers With Slow Bat Speed Against: Every Pitch Perceived as faster by batter except curve, slider, sinker")
```
```{r}
fasthoriz <- fastavgagainstall %>%
  group_by(pitch_name) %>%
  summarize(meanhorizfast = mean(pfx_x, na.rm = TRUE))

slowhoriz <- slowavgagainstall %>%
  group_by(pitch_name) %>%
  summarize(meanhorizslow = mean(pfx_x, na.rm = TRUE))

allhoriz <- fasthoriz %>%
  full_join(slowhoriz)

allhoriz <- allhoriz %>%
  rename(slow = meanhorizslow, fast = meanhorizfast)

allhoriz2<- allhoriz %>%
  pivot_longer(cols = c(slow, fast), 
               names_to = "BatSpeedAgainst", 
               values_to = "HorizBreak")

allhoriz2 %>%
  ggplot(aes(x = pitch_name, y = HorizBreak, fill = BatSpeedAgainst)) +
  geom_col(position = "dodge") +
  coord_flip() +
  theme_minimal() +
  labs(x = "Pitch Type", y = "Horizontal Break (inches)", title = "Mean Horizontal Break Comparison by Pitch Type", fill = "Bat Speed Against", caption = "Takeaway: Pitchers With Slow Bat Speed Against: More Horiz Break On Every Pitch Except Sweeper, Splitter")
```

```{r}
fastvert <- fastavgagainstall %>%
  group_by(pitch_name) %>%
  summarize(meanvertfast = mean(pfx_z, na.rm = TRUE))

slowvert <- slowavgagainstall %>%
  group_by(pitch_name) %>%
  summarize(meanvertslow = mean(pfx_z, na.rm = TRUE))

allvert <- fastvert %>%
  full_join(slowvert)

allvert <- allvert %>%
  rename(slow = meanvertslow, fast = meanvertfast)

allvert2<- allvert %>%
  pivot_longer(cols = c(slow, fast), 
               names_to = "BatSpeedAgainst", 
               values_to = "VertBreak")

allvert2 %>%
  ggplot(aes(x = pitch_name, y = VertBreak, fill = BatSpeedAgainst)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(x = "Pitch Type", y = "Vertical Break (inches)", title = "Mean Vertical Break Comparison by Pitch Type", fill = "Bat Speed Against", caption = "Takeaway: Pitchers With Slow Bat Speed Against: Less Vertical Break on Every Pitch Except Splitter")
```

# Do the above with like 5 specific pitchers to show evidence of variation (could also do boxplots)

# This is one visualization we could present

```{r}
all50against <- fastavgagainst %>%
  full_join(slowavgagainst)

all50againstall <- baseballdatacleanqual %>%
  left_join(all50against) %>%
  filter(avgbatspeeddiff != 'NA')

all50against <- all50against %>%
  mutate(PLAYERNAME = fct_reorder(PLAYERNAME, avgbatspeeddiff))

baseballdatacleanqual %>%
  ungroup() %>%
  summarize(mean(batspeeddiff))

all50against %>%
  ggplot(aes(x = PLAYERNAME, y = avgbatspeeddiff)) +
  geom_col(aes(fill = avgbatspeeddiff > 0)) +
  geom_hline(yintercept = -6.882775e-17, color = 'blue', size = 1.5) +
  labs(x = "Pitcher Name", y = "Avg Speed Difference", title = "Average Bat Speed Difference Against Compared to Batter Mean Bat Speed", caption = "Blue line Indicates average bat speed differential") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_manual(values = c("red", "darkgreen"), name = "Bat Speed Difference", labels = c("Below Average", "Above Average"))
```

```{r}
all50againstall3balls <- all50againstall %>%
  filter(balls == 3 & strikes < 2)

all50againstall3balls <- all50againstall3balls %>%
  mutate(PLAYERNAME = fct_reorder(PLAYERNAME, avgbatspeeddiff))

all50againstall3balls%>%
  ungroup() %>%
  summarize(mean(avgbatspeeddiff))

all50againstall3balls %>%
  ggplot(aes(x = PLAYERNAME, y = avgbatspeeddiff)) +
  geom_col(aes(fill = avgbatspeeddiff > 0)) +
  geom_hline(yintercept =  0.2149666, color = 'blue', size = 1.5) +
  labs(x = "Pitcher Name", y = "Avg Speed Difference", title = "Average Bat Speed Difference Against Compared to Batter Mean Bat Speed In 3 Ball Counts", caption = "Blue line Indicates average bat speed differential") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_manual(values = c("red", "darkgreen"), name = "Bat Speed Difference", labels = c("Below Average", "Above Average"))
```
```{r}
all50againstall2strikes <- all50againstall %>%
  filter(balls < 2 & strikes == 2) %>%
  ungroup()
  
all50againstall2strikes <- all50againstall2strikes %>%
  mutate(PLAYERNAME = fct_reorder(PLAYERNAME, avgbatspeeddiff))
  
all50againstall2strikes %>%
  ungroup() %>%
  summarize(mean(avgbatspeeddiff))

all50againstall2strikes %>%
  ggplot(aes(x = PLAYERNAME, y = avgbatspeeddiff)) +
  geom_col(aes(fill = avgbatspeeddiff > 0)) +
  geom_hline(yintercept = -0.3192937, color = 'blue', size = 1.5) +
  labs(x = "Pitcher Name", y = "Avg Speed Difference", title = "Average Bat Speed Difference Against Compared to Batter Mean Bat Speed In 2 Strike Counts", caption = "Blue line Indicates average bat speed differential") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_manual(values = c("red", "darkgreen"), name = "Bat Speed Difference", labels = c("Below Average", "Above Average"))
```



```{r}
vars <- c("pfx_x","pfx_z","release_speed","speeddiff","release_spin_rate")
# Overall Averages
baseballdatacleanqual %>%
  ungroup() %>%
  group_by(pitch_name) %>%
  summarize(across(vars, ~mean(. , na.rm = TRUE))) %>%
  filter(pitch_name == "4 Seam" | pitch_name == "Changeup" | pitch_name == "Curveball" | pitch_name == "K Curve" | pitch_name == "Sinker" | pitch_name == "Slider" | pitch_name == "Sweeper" | pitch_name == "Cutter" | pitch_name == "Splitter")

# Averages When Hitters Swing SLOWER than their average swing speed
baseballdatacleanqual %>%
  filter(fasterthanavg == FALSE) %>%
  ungroup() %>%
  group_by(pitch_name) %>%
  summarize(across(vars, ~mean(. , na.rm = TRUE))) %>%
  filter(pitch_name == "4 Seam" | pitch_name == "Changeup" | pitch_name == "Curveball" | pitch_name == "K Curve" | pitch_name == "Sinker" | pitch_name == "Slider" | pitch_name == "Sweeper" | pitch_name == "Cutter" | pitch_name == "Splitter")

# Averages When Hitters Swing FASTER than their average swing speed
baseballdatacleanqual %>%
  filter(fasterthanavg == TRUE) %>%
  ungroup() %>%
  group_by(pitch_name) %>%
  summarize(across(vars, ~mean(. , na.rm = TRUE))) %>%
  filter(pitch_name == "4 Seam" | pitch_name == "Changeup" | pitch_name == "Curveball" | pitch_name == "K Curve" | pitch_name == "Sinker" | pitch_name == "Slider" | pitch_name == "Sweeper" | pitch_name == "Cutter" | pitch_name == "Splitter")
```

```{r}
library(GGally)
ggpairs(pitcherdata, columns = c("fastball_avg_speed", "fastball_avg_spin", "fastball_avg_break_x", "fastball_avg_break_z","n_fastball_formatted", "avg_swing_speed"))
```

```{r}
slowerpitchcount <- baseballdatacleanqual %>%
  filter(fasterthanavg == FALSE) %>%
  ungroup() %>%
  group_by(pitch_name) %>%
  summarize(slowerbatspeed = n())

totalpitchcount <- baseballdatacleanqual %>%
  ungroup() %>%
  group_by(pitch_name) %>%
  summarize(totalpitches = n())

slowerpitchcount %>%
  full_join(totalpitchcount) %>%
  filter(pitch_name != "Forkball" & pitch_name != "Knuckleball" & pitch_name != "Slurve" & pitch_name != "Slow Curve") %>%
  mutate(propslowerbat = slowerbatspeed / totalpitches) %>%
  arrange(desc(propslowerbat))
```


```{r}
threeballs <- baseballdatacleanqual %>%
  filter(balls == 3 & strikes < 2)

twostrikes <- baseballdatacleanqual %>%
  filter(balls < 1 & strikes == 2)

threeballs %>%
  ungroup() %>%
  summarize(mean(bat_speed))

twostrikes %>%
  ungroup() %>%
  summarize(mean(bat_speed))

baseballdatacleanqual %>%
  ungroup() %>%
  summarize(mean(bat_speed))
  
```
```{r}

# Swing Slower on 3 Ball Count
threeballs %>%
 filter(fasterthanavg == FALSE) %>%
  ungroup() %>%
  group_by(pitch_name) %>%
  summarize(across(vars, ~mean(. , na.rm = TRUE))) %>%
  filter(pitch_name == "4 Seam" | pitch_name == "Changeup" | pitch_name == "Curveball" | pitch_name == "K Curve" | pitch_name == "Sinker" | pitch_name == "Slider" | pitch_name == "Sweeper" | pitch_name == "Cutter" | pitch_name == "Splitter")

# Swing faster on 2 strike count
twostrikes %>%
  filter(fasterthanavg == TRUE) %>%
  ungroup() %>%
  group_by(pitch_name) %>%
  summarize(across(vars, ~mean(. , na.rm = TRUE))) %>%
  filter(pitch_name == "4 Seam" | pitch_name == "Changeup" | pitch_name == "Curveball" | pitch_name == "K Curve" | pitch_name == "Sinker" | pitch_name == "Slider" | pitch_name == "Sweeper" | pitch_name == "Cutter" | pitch_name == "Splitter")

# Overall averages across pitch type
baseballdatacleanqual %>%
  ungroup() %>%
  group_by(pitch_name) %>%
  summarize(across(vars, ~mean(. , na.rm = TRUE))) %>%
  filter(pitch_name == "4 Seam" | pitch_name == "Changeup" | pitch_name == "Curveball" | pitch_name == "K Curve" | pitch_name == "Sinker" | pitch_name == "Slider" | pitch_name == "Sweeper" | pitch_name == "Cutter" | pitch_name == "Splitter")
```

```{r}
threeballs %>%
  ungroup() %>%
  summarize(mean(bat_speed))


threeballs %>%
  filter(pitch_name != 'Slow Curve' & pitch_name != 'Forkball' & pitch_name != "Slurve" & pitch_name != "Knuckleball") %>%
  filter(bat_speed > 50) %>%
  ggplot(aes(x = pitch_name, y = bat_speed, fill = pitch_name)) +
  geom_boxplot() +
  geom_hline(yintercept = 72.72109, color = "blue") +
  theme_minimal() +
  labs(x = "Pitch Name", 
       y = "Swing Speed", 
       title = "Distribution of Batters' Swing Speeds by Pitch Type In 3 Ball Counts") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
twostrikes %>%
  ungroup() %>%
  summarize(mean(bat_speed))


twostrikes %>%
  filter(pitch_name != 'Slow Curve' & pitch_name != 'Forkball' & pitch_name != "Slurve" & pitch_name != "Knuckleball") %>%
  filter(bat_speed > 57 & bat_speed < 78) %>%
  ggplot(aes(x = pitch_name, y = bat_speed, fill = pitch_name)) +
  geom_boxplot() +
  geom_hline(yintercept = 67.857, color = "blue") +
  theme_minimal() +
  labs(x = "Pitch Name", 
       y = "Swing Speed", 
       title = "Distribution of Batters' Swing Speeds by Pitch Type In 2 Strike Counts") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# Data Context

Our data was collected using Statcast, by CSAS (Connecticut Sports Analytics) to analyze pitcher-batter interactions in MLB. Every MLB ballpark is equipped with Statcast tracking technology, which is used to acquire the data. See this link [statcastlink](https://www.mlb.com/glossary/statcast) for more background on Statcast. It contains data on 346,250 Major League Baseball plate appearances from 4/2/2024 to 6/30/2024 (first half of the 2024 regular season), as well as 94 different variables.

Link to player ID data: [playerdatalink](https://www.smartfantasybaseball.com/tools/)

Cumulative season data created through custom Baseball Savant Leaderboard: [savantlink](https://baseballsavant.mlb.com/leaderboard/custom?year=2024&type=pitcher&filter=&min=250&selections=pa%2Ck_percent%2Cp_era%2Cavg_swing_speed%2Cfast_swing_rate%2Cavg_swing_length%2Cswords%2Cexit_velocity_avg%2Cbarrel_batted_rate%2Chard_hit_percent%2Cout_zone_percent%2Cin_zone_percent%2Cwhiff_percent%2Cf_strike_percent%2Carm_angle%2Cn_fastball_formatted%2Cfastball_avg_speed%2Cfastball_avg_spin%2Cfastball_avg_break_x%2Cfastball_avg_break_z%2Cfastball_avg_break%2Cn_breaking_formatted%2Cbreaking_avg_speed%2Cbreaking_avg_spin%2Cbreaking_avg_break_x%2Cbreaking_avg_break_z%2Cbreaking_avg_break%2Cn_offspeed_formatted%2Coffspeed_avg_speed%2Coffspeed_avg_spin%2Coffspeed_avg_break_x%2Coffspeed_avg_break_z%2Coffspeed_avg_break&chart=false&x=pa&y=pa&r=no&chartType=beeswarm&sort=pa&sortDir=asc)

## Data Codebook

The codebook for our data can be found at this link: [codebooklink](https://baseballsavant.mlb.com/csv-docs)

# Project Topic

The effect of pitchers on batters' swing speeds and swing lengths

# Questions We Wish To Explore

1.  Do pitchers modify batters' behavior? That is do batters' swing speeds differ depending on something pitchers do?

2.  Do some pitchers elicit different swing speeds and swing lengths than others? That is, are some pitchers better at slowing down batters' swing speeds or making their swings longer?

3.  If pitchers do affect a batter's swing speed/length, what is the cause of this? Is it pitch speed, pitch movement, or something else?

# Plan to Further Explore Questions

We plan to improve the visualizations we have already created, as well as create more visualizations that explore the batter/pitcher interaction. We found another dataset as suggested in our Milestone 1 Feedback, and we would like to better integrate this additional dataset into our analysis. We would also like to have a concrete answer to our third broad question, and begin interpreting our results.



```{r}
sessionInfo()
```

